---
title: 'Alpha diversity: Linear Mixed Model'
output: html_notebook
author: "Kai Wang"
---

Based on the experimental design (3 x 3 Latin Squares), we will fit linear mixed models for alpha diversity indices to test interaction effect between phase and diet type. Random effect of cow will be removed if a mixed model is singular.

```{r message=FALSE, warning=FALSE}
# loading required packages
library(lme4)
library(lmerTest)
library(tidyverse)
library(emmeans)
library(ggpubr)
```



# 1. Extract data from Qiime2 artifact files

## 1.1 Bacteria

```{r message=FALSE, warning=FALSE}
# observed features
system('qiime tools export \\
          --input-path Results/2-core-metrics_bac/observed_features_vector.qza \\
          --output-path Results/2-core-metrics_bac/')
system('mv Results/2-core-metrics_bac/alpha-diversity.tsv Results/2-core-metrics_bac/observed_features_vector.tsv')

# faith pd
system('qiime tools export \\
          --input-path Results/2-core-metrics_bac/faith_pd_vector.qza \\
          --output-path Results/2-core-metrics_bac/')
system('mv Results/2-core-metrics_bac/alpha-diversity.tsv Results/2-core-metrics_bac/faith_pd_vector.tsv')

# shannon
system('qiime tools export \\
          --input-path Results/2-core-metrics_bac/shannon_vector.qza \\
          --output-path Results/2-core-metrics_bac/')
system('mv Results/2-core-metrics_bac/alpha-diversity.tsv Results/2-core-metrics_bac/shannon_vector.tsv')

# eveness
system('qiime tools export \\
          --input-path Results/2-core-metrics_bac/evenness_vector.qza \\
          --output-path Results/2-core-metrics_bac/')
system('mv Results/2-core-metrics_bac/alpha-diversity.tsv Results/2-core-metrics_bac/evenness_vector.tsv')
```

## 1.2 Archaea

```{r message=FALSE, warning=FALSE}
# observed features
system('qiime tools export \\
          --input-path Results/2-core-metrics_arc/observed_features_vector.qza \\
          --output-path Results/2-core-metrics_arc/')
system('mv Results/2-core-metrics_arc/alpha-diversity.tsv Results/2-core-metrics_arc/observed_features_vector.tsv')

# faith pd
system('qiime tools export \\
          --input-path Results/2-core-metrics_arc/faith_pd_vector.qza \\
          --output-path Results/2-core-metrics_arc/')
system('mv Results/2-core-metrics_arc/alpha-diversity.tsv Results/2-core-metrics_arc/faith_pd_vector.tsv')

# shannon
system('qiime tools export \\
          --input-path Results/2-core-metrics_arc/shannon_vector.qza \\
          --output-path Results/2-core-metrics_arc/')
system('mv Results/2-core-metrics_arc/alpha-diversity.tsv Results/2-core-metrics_arc/shannon_vector.tsv')

# eveness
system('qiime tools export \\
          --input-path Results/2-core-metrics_arc/evenness_vector.qza \\
          --output-path Results/2-core-metrics_arc/')
system('mv Results/2-core-metrics_arc/alpha-diversity.tsv Results/2-core-metrics_arc/evenness_vector.tsv')
```


# 2. Data preparation

## 2.1 Loading metrics and metadata

```{r}
# load and combine bacteria metrics into a data frame
files_bac <- list.files('Results/2-core-metrics_bac', "*vector.tsv", recursive = TRUE, full.names = TRUE)

data_bac <- lapply(files_bac, function(x){read.delim(x, row.names = 1)})
data_bac <- do.call(cbind, data_bac)
```

```{r}
# load and combine archaea metrics into a data frame
files_arc <- list.files('Results/2-core-metrics_arc', "*vector.tsv", recursive = TRUE, full.names = TRUE)

data_arc <- lapply(files_arc, function(x){read.delim(x, row.names = 1)})
data_arc <- do.call(cbind, data_arc)
```

```{r}
# metadata
metadata <- read.delim('Data/0-exported-metadata/sra-metadata.tsv', row.names = 1, check.names = FALSE)
```

## 2.2 Curate metadata

Column `Title` in the metadata contains information about animal identification and PCR library types, which are retrieved here. A and B stand for corresponding archaea or bacteria library.

```{r}
metadata[, 'Animal_ID'] <- str_extract(string = metadata$Title, pattern = '[0-9][0-9]')
metadata[, 'PCR'] <- ifelse(str_detect(metadata$Title, 'A$'), 'A', 'B')
```

Column `Description [sample]` contains information about diet treatments [Diet], and sample types [Phase] from which parts of rumen contents samples were collected. However, period of treatment cannot be inferred here with a high confidence, random effect of period will not be considered in linear mixed models.

```{r}
metadata[, 'Phase'] <- do.call(paste,
                               data.frame(
                                 str_split_fixed(
                                   string = metadata$`Description [sample]`, 
                                   pattern = ' ', n = 3)[, 1:2]))
metadata[, 'Diet'] <- str_split_fixed(string = metadata$`Description [sample]`, pattern = ' ', n = 3)[, 3]

metadata$Phase <- factor(metadata$Phase, levels = c("solid phase", "liquid phase", "rumen fluid"), 
                         labels = c("Solid phase", "Liquid phase", "Rumen fluid"))

metadata$Diet <- factor(metadata$Diet, levels = c("corn", "grass", "hay"), 
                         labels = c("Corn silage", "Grass silage", "Grass hay"))
```

```{r}
metadata <- metadata[c('Animal_ID', 'PCR', 'Phase', 'Diet')]
metadata[, 'Animal_ID'] <- as.factor(metadata[, 'Animal_ID'])
head(metadata)
```

## 2.3 Merging alpha diversity indices and metadata

```{r}
data_arc_merged <- cbind(data_arc, metadata[rownames(data_arc), ])
data_bac_merged <- cbind(data_bac, metadata[rownames(data_bac), ])
```

# 3. Model fitting

## 3.1 Bacteria

### model 

```{r}
# observed features
fit_obsf_bac <- lmer(observed_features ~ Phase * Diet + (1|Animal_ID),
     data = data_bac_merged)
#faith pd
fit_fapd_bac <- lmer(faith_pd ~ Phase * Diet + (1|Animal_ID),
     data = data_bac_merged)
# shannon
fit_shan_bac <- lmer(shannon_entropy ~ Phase * Diet + (1|Animal_ID),
     data = data_bac_merged)
# pielou's evenness
fit_even_bac <- lmer(pielou_evenness ~ Phase * Diet + (1|Animal_ID),
     data = data_bac_merged)
```

### model diagnosis

```{r}
qqnorm(residuals(fit_obsf_bac))
qqline(residuals(fit_obsf_bac))

qqnorm(residuals(fit_fapd_bac))
qqline(residuals(fit_fapd_bac))

qqnorm(residuals(fit_shan_bac))
qqline(residuals(fit_shan_bac))

qqnorm(residuals(fit_even_bac))
qqline(residuals(fit_even_bac))
```

### fixed effects, ANOVA

```{r}
anova(fit_obsf_bac)
anova(fit_fapd_bac)
anova(fit_shan_bac)
anova(fit_even_bac)
```

### fixed effects, LSM

```{r}
# observed features
emmeans(fit_obsf_bac, ~ Diet)
emmeans(fit_obsf_bac, ~ Phase)
```
```{r}
# faith pd
emmeans(fit_fapd_bac, ~ Diet)
emmeans(fit_fapd_bac, ~ Phase)
```
```{r}
# shannon
emmeans(fit_shan_bac, ~ Diet)
emmeans(fit_shan_bac, ~ Phase)
```

```{r}
# shannon
emmeans(fit_even_bac, ~ Diet)
emmeans(fit_even_bac, ~ Phase)
```

## 3.2 Archaea

### model 

```{r}
# observed features
fit_obsf_arc <- lmer(observed_features ~ Phase * Diet + (1|Animal_ID),
     data = data_arc_merged)
#faith pd
fit_fapd_arc <- lmer(faith_pd ~ Phase * Diet + (1|Animal_ID),
     data = data_arc_merged)
# shannon
fit_shan_arc <- lmer(shannon_entropy ~ Phase * Diet + (1|Animal_ID),
     data = data_arc_merged)
# pielou's evenness
fit_even_arc <- lmer(pielou_evenness ~ Phase * Diet + (1|Animal_ID),
     data = data_arc_merged)
```

### model diagnosis

```{r}
qqnorm(residuals(fit_obsf_arc))
qqline(residuals(fit_obsf_arc))

qqnorm(residuals(fit_fapd_arc))
qqline(residuals(fit_fapd_arc))

qqnorm(residuals(fit_shan_arc))
qqline(residuals(fit_shan_arc))

qqnorm(residuals(fit_even_arc))
qqline(residuals(fit_even_arc))
```

### fixed effects, ANOVA

```{r}
anova(fit_obsf_arc)
anova(fit_fapd_arc)
anova(fit_shan_arc)
anova(fit_even_arc)
```

### fixed effects, LSM

```{r}
# observed features
emmeans(fit_obsf_arc, ~ Diet)
emmeans(fit_obsf_arc, ~ Phase)
```
```{r}
# faith pd
emmeans(fit_fapd_arc, ~ Diet)
emmeans(fit_fapd_arc, ~ Phase)
```
```{r}
# shannon
emmeans(fit_shan_arc, ~ Diet)
emmeans(fit_shan_arc, ~ Phase)
```

```{r}
# shannon
emmeans(fit_even_arc, ~ Diet)
emmeans(fit_even_arc, ~ Phase)
```

# 4 Visualization

## 4.1 Defining visualize function

```{r}

# vis <- function(fit, ylab = "", xadj = 0.25, yadj = 0.05){
#   
#   # LSM
#   lsm_df <- data.frame(emmeans(fit, ~ Diet | Phase))
#   
#   # contrast
#   contrast_df <- data.frame(contrast(emmeans(fit, ~ Diet | Phase), 'pairwise'))
#   contrast_df[, c('start', 'end')] <- str_split_fixed(contrast_df$contrast, pattern = " - ", n = 2)
#   contrast_levels <- union(unique(contrast_df$start), unique(contrast_df$end))
#   contrast_df[, 'contrast_id'] <- contrast_df[, 'contrast']
#   for (i in 1:length(contrast_levels)) {
#       contrast_df[, 'contrast_id'] <- gsub(pattern = contrast_levels[i], 
#                                            replacement = i, 
#                                            x = contrast_df[, 'contrast_id'])
#   }
#   
#   contrast_df[, 'x'] <- ifelse(contrast_df$Phase == "Solid phase", 1,
#                                ifelse(contrast_df$Phase == "Liquid phase", 2, 3))
#   
#   contrast_df[, c('xmin')] <- ifelse(contrast_df$start == "Corn silage", -xadj, 
#                                      ifelse(contrast_df$start == "Grass Silage", 0, +xadj))
#   contrast_df[, c('xmax')] <- ifelse(contrast_df$end == "Corn silage", -xadj, 
#                                      ifelse(contrast_df$end == "Grass silage", 0, +xadj))
#   contrast_df[, c('xmin')] <-  contrast_df[, c('xmin')] + contrast_df$x
#   contrast_df[, c('xmax')] <-  contrast_df[, c('xmax')] + contrast_df$x
#   
#   
#   for (i in 1:nrow(contrast_df)) {
#     idx = which(lsm_df$Phase == contrast_df$Phase[i] & lsm_df$Diet == contrast_df$start[i])
#     contrast_df[i, 'y_start'] <- (lsm_df[idx, 'emmean'] + lsm_df[idx, 'SE'])*(1+yadj)
#   }
#   
#   contrast_df[, 'y_middle'] <- rep(contrast_df[contrast_df$start == "grass", 'y_start'], each = 3)
#   
#   contrast_df[, 'y'] <- ifelse(contrast_df$estimate > 0, 
#                                contrast_df[, 'y_start'], 
#                                contrast_df[, 'y_start']-contrast_df[, 'estimate'])
#   
#   contrast_df[, 'y'] <- ifelse(contrast_df$y < contrast_df$y_middle,
#                                contrast_df$y_middle,
#                                contrast_df$y)
#   
#   contrast_df[, 'y_1_vs_2'] <- rep(contrast_df[contrast_df$contrast_id == "1 - 2", 'y'], 
#                                      each = 3)
#   contrast_df[, 'y_2_vs_3'] <- rep(contrast_df[contrast_df$contrast_id == "2 - 3", 'y'], 
#                                      each = 3)
#   
#   contrast_df[, 'y'] <- ifelse(contrast_df$contrast_id == "1 - 3",
#                                ifelse(contrast_df$y >= contrast_df$y_1_vs_2*(1+yadj), 
#                                       contrast_df$y, 
#                                       contrast_df$y_1_vs_2*(1+yadj)), 
#                                contrast_df$y
#                                )
#     contrast_df[, 'y'] <- ifelse(contrast_df$contrast_id == "1 - 3",
#                                ifelse(contrast_df$y >= contrast_df$y_2_vs_3*(1+yadj), 
#                                       contrast_df$y, 
#                                       contrast_df$y_2_vs_3*(1+yadj)), 
#                                contrast_df$y
#                                )
# 
#   
#   
#   contrast_df[, 'p'] <- "*"
#   contrast_df <- contrast_df[contrast_df$p.value < 0.05, ]
#   
# 
#   lsm_df$Phase <- factor(lsm_df$Phase, 
#                          levels = c('Solid phase', 'Liquid phase', 'Rumen fluid'),
#                          labels = c('Solid\nphase', 'Liquid\nphase', 'Rumen\nfluid'))
#   
#   lsm_df$Diet <- factor(lsm_df$Diet, 
#                          levels = c('Corn silage', 'Grass silage', 'Grass hay'),
#                          labels = c('Corn silage', 'Grass silage', 'Grass hay'))
#   
#   # plot
#   plot0 <- ggplot(lsm_df, aes(Phase, emmean, fill = Diet, color = Diet)) +
#             geom_col(position = position_dodge2()) +
#             geom_errorbar(aes(ymin=emmean-SE, ymax=emmean+SE), 
#                           position = position_dodge(width = 0.9),
#                           width = 0.5, size = 0.75) +
#             labs(y = ylab) +
#             theme(plot.background = element_blank(),
#                   panel.grid = element_blank(),
#                   panel.background = element_blank(),
#                   axis.line = element_line(linewidth = 1, colour = 'black'),
#                   axis.title = element_text(size = 20, face = "bold", colour = "black"),
#                   axis.text = element_text(size = 18, face = "bold", colour = "black"),
#                   axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5),
#                   axis.title.x = element_text(margin = unit(c(5, 0, 0, 0), 'mm')),
#                   axis.title.y = element_text(margin = unit(c(0, 5, 0, 0), 'mm')),
#                   legend.text = element_text(size = 18, face = "bold", colour = "black"),
#                   legend.title = element_text(size = 18, face = "bold", colour = "black")
#                   ) +
#     scale_fill_nejm() +
#     scale_color_nejm()
#   
#   if (nrow(contrast_df) > 0) {
#     plot0 <- plot0 + 
#       geom_signif(annotations = contrast_df$p, 
#                   xmin = contrast_df$xmin,
#                   xmax = contrast_df$xmax,
#                   y_position =contrast_df$y, 
#                   size = 0.75, vjust = 0.5, tip_length = 0.05,
#                   color="black")
#   }
#   
# 
#   return(plot0)
# }

```

```{r}
myboxplot <- function(alpha_vector_table, index = "", ylab = ""){
  plot0 <- ggplot(alpha_vector_table, aes_string("Phase", index, color = "Phase")) +
    geom_boxplot() +
    geom_point() +
    facet_grid(~Diet) +
    scale_color_nejm() +
    labs(y = ylab) +
    theme_bw() +
    theme(panel.grid = element_blank(),
        axis.ticks.x = element_blank(),
        axis.line = element_line(linewidth = 1, colour = 'black'),
        axis.text.x = element_blank(),
        axis.title.y = element_text(size = 20, face = "bold", colour = "black"),
        axis.text.y = element_text(size = 18, face = "bold", colour = "black"),
        axis.title.x.bottom = element_blank(),
        axis.title.y.left = element_text(margin = unit(c(0, 5, 0, 0), 'mm')),
        legend.text = element_text(size = 18, face = "bold", colour = "black"),
        legend.title = element_text(size = 18, face = "bold", colour = "black"),
        strip.text = element_text(size = 18, face = "bold", colour = "black"),
        legend.position = "bottom"
        )
  return(plot0)
}
```

```{r}
myboxplot(data_bac_merged, "observed_features", "Observed features")
anova(fit_obsf_bac)
```
```{r}
myboxplot(data_bac_merged, "faith_pd", "Faith PD")
ggsave("Results/2-alpha_boxplot_faith_pd_bac.png", width = 9, height = 6)
anova(fit_fapd_bac)
```

```{r}
myboxplot(data_arc_merged, "observed_features", "Observed features")
anova(fit_obsf_arc)
```

```{r}
myboxplot(data_arc_merged, "faith_pd", "Faith PD")
ggsave("Results/2-alpha_boxplot_faith_pd_arc.png", width = 9, height = 6)
anova(fit_fapd_arc)
```
