---
title: "2-Beta_diversity_adonis"
author: "Kai Wang"
date: "2022-12-19"
output:
  html_document: default
  pdf_document: default
---

```{r message=FALSE, warning=FALSE}
# loading required packages
library(tidyverse)
library(vegan)
```

# 1. Extract data from Qiime2 artifact files

## 1.1 Bacteria

```{r}
'Results/2-core-metrics_bac/bray_curtis_distance_matrix.qza'
'Results/2-core-metrics_bac/unweighted_unifrac_distance_matrix.qza'
'Results/2-core-metrics_bac/weighted_unifrac_distance_matrix.qza'
'Results/2-core-metrics_bac/jaccard_distance_matrix.qza'
```

```{r message=FALSE, warning=FALSE}
# bray curtis
system('qiime tools export \\
          --input-path Results/2-core-metrics_bac/bray_curtis_distance_matrix.qza \\
          --output-path Results/2-core-metrics_bac/')
system('mv Results/2-core-metrics_bac/distance-matrix.tsv Results/2-core-metrics_bac/bray_curtis_matrix.tsv')

# jaccard
system('qiime tools export \\
          --input-path Results/2-core-metrics_bac/jaccard_distance_matrix.qza \\
          --output-path Results/2-core-metrics_bac/')
system('mv Results/2-core-metrics_bac/distance-matrix.tsv Results/2-core-metrics_bac/jaccard_matrix.tsv')

# weighted unifrac
system('qiime tools export \\
          --input-path Results/2-core-metrics_bac/weighted_unifrac_distance_matrix.qza \\
          --output-path Results/2-core-metrics_bac/')
system('mv Results/2-core-metrics_bac/distance-matrix.tsv Results/2-core-metrics_bac/weighted_unifrac_matrix.tsv')

# unweighted unifrac
system('qiime tools export \\
          --input-path Results/2-core-metrics_bac/unweighted_unifrac_distance_matrix.qza \\
          --output-path Results/2-core-metrics_bac/')
system('mv Results/2-core-metrics_bac/distance-matrix.tsv Results/2-core-metrics_bac/unweighted_unifrac_matrix.tsv')
```

## 1.2 Archaea

```{r message=FALSE, warning=FALSE}
# bray curtis
system('qiime tools export \\
          --input-path Results/2-core-metrics_arc/bray_curtis_distance_matrix.qza \\
          --output-path Results/2-core-metrics_arc/')
system('mv Results/2-core-metrics_arc/distance-matrix.tsv Results/2-core-metrics_arc/bray_curtis_matrix.tsv')

# jaccard
system('qiime tools export \\
          --input-path Results/2-core-metrics_arc/jaccard_distance_matrix.qza \\
          --output-path Results/2-core-metrics_arc/')
system('mv Results/2-core-metrics_arc/distance-matrix.tsv Results/2-core-metrics_arc/jaccard_matrix.tsv')

# weighted unifrac
system('qiime tools export \\
          --input-path Results/2-core-metrics_arc/weighted_unifrac_distance_matrix.qza \\
          --output-path Results/2-core-metrics_arc/')
system('mv Results/2-core-metrics_arc/distance-matrix.tsv Results/2-core-metrics_arc/weighted_unifrac_matrix.tsv')

# unweighted unifrac
system('qiime tools export \\
          --input-path Results/2-core-metrics_arc/unweighted_unifrac_distance_matrix.qza \\
          --output-path Results/2-core-metrics_arc/')
system('mv Results/2-core-metrics_arc/distance-matrix.tsv Results/2-core-metrics_arc/unweighted_unifrac_matrix.tsv')
```


# 2. Data preparation

## 2.1 Loading metrics and metadata

```{r}
# load and combine bacteria metrics into a data frame
files_bac <- list.files('Results/2-core-metrics_bac', "*matrix.tsv", recursive = TRUE, full.names = TRUE)
data_bac <- lapply(files_bac, function(x){read.delim(x, row.names = 1)})
names(data_bac) <- c("bray_curtis_matrix", "jaccard", "unweighted_unifrac", "weighted_unifrac")
```

```{r}
# load and combine archaea metrics into a data frame
files_arc <- list.files('Results/2-core-metrics_arc', "*matrix.tsv", recursive = TRUE, full.names = TRUE)
data_arc <- lapply(files_arc, function(x){read.delim(x, row.names = 1)})
names(data_arc) <- c("bray_curtis_matrix", "jaccard", "unweighted_unifrac", "weighted_unifrac")
```

```{r}
# metadata
metadata <- read.delim('Data/0-exported-metadata/sra-metadata.tsv', row.names = 1, check.names = FALSE)
```

## 2.2 Curate metadata

Column `Title` in the metadata contains information about animal identification and PCR library types, which are retrieved here. A and B stand for corresponding archaea or bacteria library.

```{r}
metadata[, 'Animal_ID'] <- str_extract(string = metadata$Title, pattern = '[0-9][0-9]')
metadata[, 'PCR'] <- ifelse(str_detect(metadata$Title, 'A$'), 'A', 'B')
```

Column `Description [sample]` contains information about diet treatments [Diet], and sample types [Phase] from which parts of rumen contents samples were collected. However, period of treatment cannot be inferred here with a high confidence, random effect of period will not be considered in linear mixed models.

```{r}
metadata[, 'Phase'] <- do.call(paste,
                               data.frame(
                                 str_split_fixed(
                                   string = metadata$`Description [sample]`, 
                                   pattern = ' ', n = 3)[, 1:2]))
metadata[, 'Diet'] <- str_split_fixed(string = metadata$`Description [sample]`, pattern = ' ', n = 3)[, 3]
```

```{r}
metadata <- metadata[c('Animal_ID', 'PCR', 'Phase', 'Diet')]
metadata[, 'Animal_ID'] <- as.factor(metadata[, 'Animal_ID'])
head(metadata)
```

# 3. Visualization, scatter plot

# 3.1 Function

```{r message=FALSE, warning=FALSE}
vis <- function(distance_matrix, map, color = "", shape = "", point_size = 2){
  # color = "Phase"
  # shape = "Diet"
  # distance_matrix <- data_arc$bray_curtis_matrix
  # point_size = 2

  pcoa <- cmdscale(distance_matrix, eig = TRUE, k = 4)
  coordinates <- data.frame(pcoa$points)
  eig_percentage <- round(pcoa$eig/sum(pcoa$eig) * 100, 1)

  coordinates <- cbind(coordinates, metadata[rownames(coordinates), ])


  plot0 <- ggplot(coordinates, aes_string("X1", "X2", group = color, color = color, shape = shape)) +
    geom_point(size = point_size) +
    stat_ellipse(level = 0.68) +
    labs(x = paste0("PCoA 1: ", eig_percentage[1], "%"),
        y = paste0("PCoA 2: ", eig_percentage[2], "%")) +
    theme(plot.background = element_blank(),
          panel.grid = element_blank(),
          panel.background = element_blank(),
          axis.line = element_line(linewidth = 1, colour = 'black'),
          axis.title = element_text(size = 20, face = "bold", colour = "black"),
          axis.text = element_text(size = 18, face = "bold", colour = "black"),
          axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5),
          axis.title.x = element_text(margin = unit(c(5, 0, 0, 0), 'mm')),
          axis.title.y = element_text(margin = unit(c(0, 5, 0, 0), 'mm')),
          legend.text = element_text(size = 18, face = "bold", colour = "black"),
          legend.title = element_text(size = 18, face = "bold", colour = "black"),
          legend.background = element_blank(),
          legend.key = element_blank()
          )
  return(plot0)
}
```

## 3.2 Bacteria

```{r message=FALSE, warning=FALSE}
vis(data_bac$bray_curtis_matrix, metadata, color = "Phase", shape = "Diet")
vis(data_bac$jaccard, metadata, color = "Phase", shape = "Diet")
vis(data_bac$unweighted_unifrac, metadata, color = "Phase", shape = "Diet")
vis(data_bac$weighted_unifrac, metadata, color = "Phase", shape = "Diet")
```

## 3.3 Archaea
```{r message=FALSE, warning=FALSE}
vis(data_arc$bray_curtis_matrix, metadata, color = "Phase", shape = "Diet")
vis(data_arc$jaccard, metadata, color = "Phase", shape = "Diet")
vis(data_arc$unweighted_unifrac, metadata, color = "Phase", shape = "Diet")
vis(data_arc$weighted_unifrac, metadata, color = "Phase", shape = "Diet")
```

# 4. Adonis

## 4.1 Bacteria

### metadata

```{r}
metadata_bac <- metadata[rownames(data_bac$bray_curtis_matrix), ]
```

### interaction

```{r}
adonis2(data_bac$bray_curtis_matrix ~ Phase * Diet, metadata_bac)
adonis2(data_bac$jaccard ~ Phase * Diet, metadata_bac)
adonis2(data_bac$unweighted_unifrac ~ Phase * Diet, metadata_bac)
adonis2(data_bac$weighted_unifrac ~ Phase * Diet, metadata_bac)
```

### diet effect

```{r}
with(metadata_bac, adonis2(data_bac$bray_curtis_matrix ~ Diet, metadata_bac, strata = Phase))
with(metadata_bac, adonis2(data_bac$jaccard ~ Diet, metadata_bac, strata = Phase))
with(metadata_bac, adonis2(data_bac$unweighted_unifrac ~ Diet, metadata_bac, strata = Phase))
with(metadata_bac, adonis2(data_bac$weighted_unifrac ~ Diet, metadata_bac, strata = Phase))
```




## 4.2 Archaea

### metadata

```{r}
metadata_arc <- metadata[rownames(data_arc$bray_curtis_matrix), ]
```

### interaction

```{r}
adonis2(data_arc$bray_curtis_matrix ~ Phase * Diet, metadata_arc)
adonis2(data_arc$jaccard ~ Phase * Diet, metadata_arc)
adonis2(data_arc$unweighted_unifrac ~ Phase * Diet, metadata_arc)
adonis2(data_arc$weighted_unifrac ~ Phase * Diet, metadata_arc)
```

### diet effect

```{r}
with(metadata_arc, adonis2(data_arc$bray_curtis_matrix ~ Diet, metadata_arc, strata = Phase))
with(metadata_arc, adonis2(data_arc$jaccard ~ Diet, metadata_arc, strata = Phase))
with(metadata_arc, adonis2(data_arc$unweighted_unifrac ~ Diet, metadata_arc, strata = Phase))
with(metadata_arc, adonis2(data_arc$weighted_unifrac ~ Diet, metadata_arc, strata = Phase))
```


