---
title: 'Alpha diversity: Linear Mixed Model'
output:
  html_document:
    df_print: paged
---

> Based on the experimental design (3 x 3 Latin Squares), alpha diversity indices were fitted with linear mixed models to test interaction effects between phase and diet type.

```{r message=FALSE, warning=FALSE}
# loading required packages
library(lme4)
library(lmerTest)
library(tidyverse)
library(emmeans)
library(ggpubr)
```


# 1. Export data from Qiime2 artifact files

```{r message=FALSE, warning=FALSE}
# observed features
system('qiime tools export \\
          --input-path Results/2-core-metrics_bac/observed_features_vector.qza \\
          --output-path Results/2-core-metrics_bac/')
system('mv Results/2-core-metrics_bac/alpha-diversity.tsv Results/2-core-metrics_bac/observed_features.tsv')

# faith pd
system('qiime tools export \\
          --input-path Results/2-core-metrics_bac/faith_pd_vector.qza \\
          --output-path Results/2-core-metrics_bac/')
system('mv Results/2-core-metrics_bac/alpha-diversity.tsv Results/2-core-metrics_bac/faith_pd.tsv')

# shannon
system('qiime tools export \\
          --input-path Results/2-core-metrics_bac/shannon_vector.qza \\
          --output-path Results/2-core-metrics_bac/')
system('mv Results/2-core-metrics_bac/alpha-diversity.tsv Results/2-core-metrics_bac/shannon.tsv')

# eveness
system('qiime tools export \\
          --input-path Results/2-core-metrics_bac/evenness_vector.qza \\
          --output-path Results/2-core-metrics_bac/')
system('mv Results/2-core-metrics_bac/alpha-diversity.tsv Results/2-core-metrics_bac/evenness.tsv')
```

# 2. Data preparation

## 2.1 Loading metrics and metadata

```{r}
# load data
observed <- read.delim('Results/2-core-metrics_bac/observed_features.tsv', row.names = 1)
faith_pd <- read.delim('Results/2-core-metrics_bac/faith_pd.tsv', row.names = 1)
shannon <- read.delim('Results/2-core-metrics_bac/shannon.tsv', row.names = 1)
evenness <- read.delim('Results/2-core-metrics_bac/evenness.tsv', row.names = 1)
metadata <- read.delim('Data/0-exported-metadata/sra-metadata.tsv', row.names = 1, check.names = FALSE)
```

## 2.2 Curate metadata

Column `Title` in the metadata contains information about animal identification and PCR library types, which are retrieved here. A and B stand for corresponding archaea or bacteria library.

```{r}
metadata[, 'Animal_ID'] <- str_extract(string = metadata$Title, pattern = '[0-9][0-9]')
metadata[, 'PCR'] <- ifelse(str_detect(metadata$Title, 'A$'), 'A', 'B')
```

Column `Description [sample]` contains information about diet treatments [Diet], and sample types [Phase] from which parts of rumen contents samples were collected. However, period of treatment cannot be inferred here with a high confidence, random effect of period will not be considered in linear mixed models.

```{r}
metadata[, 'Phase'] <- do.call(paste,
                               data.frame(
                                 str_split_fixed(
                                   string = metadata$`Description [sample]`, 
                                   pattern = ' ', n = 3)[, 1:2]))
metadata[, 'Diet'] <- str_split_fixed(string = metadata$`Description [sample]`, pattern = ' ', n = 3)[, 3]
```

```{r}
metadata <- metadata[c('Animal_ID', 'PCR', 'Phase', 'Diet')]
metadata[, 'Animal_ID'] <- as.factor(metadata[, 'Animal_ID'])
head(metadata)
```

## 2.3 Merging alpha diversity indices and metadata

```{r}
data_merged <- cbind(observed, faith_pd, evenness, shannon, metadata[rownames(observed), ])
```

# 3. Model fitting


## 3.1 Observed features

### model
```{r}
fit_obsf <- lmer(observed_features ~ Phase * Diet + (1|Animal_ID),
     data = data_merged)
```

### anova: fixed effects

```{r}
anova(fit_obsf)
```
### model diagnosis

```{r}
qqnorm(residuals(fit_obsf))
qqline(residuals(fit_obsf))
```
## 3.2 Faith PD

### model

Model fitting is singular, so we can leave random effect of animal out of model.

```{r}
fit_fapd <- lmer(faith_pd ~ Phase * Diet + (1|Animal_ID),
     data = data_merged)
```

### anova: fixed effects

```{r}
anova(fit_fapd)
```
### model diagnosis

```{r}
qqnorm(residuals(fit_fapd))
qqline(residuals(fit_fapd))
```
## 3.3 Shannon

### model

Model fitting is singular, so we can leave random effect of animal out of model.

```{r}
fit_shan <- lmer(shannon_entropy ~ Phase * Diet + (1|Animal_ID),
     data = data_merged)
```

### anova: fixed effects

```{r}
anova(fit_shan)
```
### model diagnosis

```{r}
qqnorm(residuals(fit_shan))
qqline(residuals(fit_shan))
```
## 3.4 evenness

### model

Model fitting is singular, so we can leave random effect of animal out of model.

```{r}
fit_even <- lmer(pielou_evenness ~ Phase * Diet + (1|Animal_ID),
     data = data_merged)
```

### anova: fixed effects

```{r}
anova(fit_even)
```
### model diagnosis

```{r}
qqnorm(residuals(fit_even))
qqline(residuals(fit_even))
```










# 4. Visulization

## 4.1 Defining visualize function

```{r}
vis <- function(fit, ylab = "", xadj = 0.25, yadj = 0.05){
  # fit = fit_obsf
  # xadj = 0.25
  # yadj = 0.1
  # ylab = ""
  
  # LSM
  lsm_df <- data.frame(emmeans(fit, ~ Diet | Phase))
  
  # contrast
  contrast_df <- data.frame(contrast(emmeans(fit, ~ Diet | Phase), 'pairwise'))
  contrast_df[, c('start', 'end')] <- str_split_fixed(contrast_df$contrast, pattern = " - ", n = 2)
  contrast_levels <- union(unique(contrast_df$start), unique(contrast_df$end))
  contrast_df[, 'contrast_id'] <- contrast_df[, 'contrast']
  for (i in 1:length(contrast_levels)) {
      contrast_df[, 'contrast_id'] <- gsub(pattern = contrast_levels[i], 
                                           replacement = i, 
                                           x = contrast_df[, 'contrast_id'])
  }
  
  contrast_df[, 'x'] <- ifelse(contrast_df$Phase == "solid phase", 1,
                               ifelse(contrast_df$Phase == "liquid phase", 2, 3))
  
  contrast_df[, c('xmin')] <- ifelse(contrast_df$start == "corn", -xadj, 
                                     ifelse(contrast_df$start == "grass", 0, +xadj))
  contrast_df[, c('xmax')] <- ifelse(contrast_df$end == "corn", -xadj, 
                                     ifelse(contrast_df$end == "grass", 0, +xadj))
  contrast_df[, c('xmin')] <-  contrast_df[, c('xmin')] + contrast_df$x
  contrast_df[, c('xmax')] <-  contrast_df[, c('xmax')] + contrast_df$x
  
  
  for (i in 1:nrow(contrast_df)) {
    idx = which(lsm_df$Phase == contrast_df$Phase[i] & lsm_df$Diet == contrast_df$start[i])
    contrast_df[i, 'y_start'] <- (lsm_df[idx, 'emmean'] + lsm_df[idx, 'SE'])*(1+yadj)
  }
  
  contrast_df[, 'y_middle'] <- rep(contrast_df[contrast_df$start == "grass", 'y_start'], each = 3)
  
  contrast_df[, 'y'] <- ifelse(contrast_df$estimate > 0, 
                               contrast_df[, 'y_start'], 
                               contrast_df[, 'y_start']-contrast_df[, 'estimate'])
  
  contrast_df[, 'y'] <- ifelse(contrast_df$y < contrast_df$y_middle,
                               contrast_df$y_middle,
                               contrast_df$y)
  
  contrast_df[, 'y_1_vs_2'] <- rep(contrast_df[contrast_df$contrast_id == "1 - 2", 'y'], 
                                     each = 3)
  contrast_df[, 'y_2_vs_3'] <- rep(contrast_df[contrast_df$contrast_id == "2 - 3", 'y'], 
                                     each = 3)
  
  contrast_df[, 'y'] <- ifelse(contrast_df$contrast_id == "1 - 3",
                               ifelse(contrast_df$y >= contrast_df$y_1_vs_2*(1+yadj), 
                                      contrast_df$y, 
                                      contrast_df$y_1_vs_2*(1+yadj)), 
                               contrast_df$y
                               )
    contrast_df[, 'y'] <- ifelse(contrast_df$contrast_id == "1 - 3",
                               ifelse(contrast_df$y >= contrast_df$y_2_vs_3*(1+yadj), 
                                      contrast_df$y, 
                                      contrast_df$y_2_vs_3*(1+yadj)), 
                               contrast_df$y
                               )

  
  
  contrast_df[, 'p'] <- "*"
  contrast_df <- contrast_df[contrast_df$p.value < 0.05, ]
  

  lsm_df$Phase <- factor(lsm_df$Phase, 
                         levels = c('solid phase', 'liquid phase', 'rumen fluid'),
                         labels = c('Solid\nphase', 'Liquid\nphase', 'Rumen\nfluid'))
  
  lsm_df$Diet <- factor(lsm_df$Diet, 
                         levels = c('corn', 'grass', 'hay'),
                         labels = c('Corn', 'Grass', 'Hay'))
  
  # plot
  plot0 <- ggplot(lsm_df, aes(Phase, emmean, fill = Diet, color = Diet)) +
            geom_col(position = position_dodge2()) +
            geom_errorbar(aes(ymin=emmean-SE, ymax=emmean+SE), 
                          position = position_dodge(width = 0.9),
                          width = 0.5, size = 0.75) +
            labs(y = ylab) +
            theme(plot.background = element_blank(),
                  panel.grid = element_blank(),
                  panel.background = element_blank(),
                  axis.line = element_line(linewidth = 1, colour = 'black'),
                  axis.title = element_text(size = 20, face = "bold", colour = "black"),
                  axis.text = element_text(size = 18, face = "bold", colour = "black"),
                  axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5),
                  axis.title.x = element_text(margin = unit(c(5, 0, 0, 0), 'mm')),
                  axis.title.y = element_text(margin = unit(c(0, 5, 0, 0), 'mm')),
                  legend.text = element_text(size = 18, face = "bold", colour = "black"),
                  legend.title = element_text(size = 18, face = "bold", colour = "black")
                  )
  
  if (nrow(contrast_df) > 0) {
    plot0 <- plot0 + 
      geom_signif(annotations = contrast_df$p, 
                  xmin = contrast_df$xmin,
                  xmax = contrast_df$xmax,
                  y_position =contrast_df$y, 
                  size = 0.75, vjust = 0.5, tip_length = 0.05,
                  color="black")
  }
  

  return(plot0)
}

```

```{r}
vis(fit_obsf, 'Observed features')
```

```{r}
vis(fit_fapd, 'Faith PD')
```

```{r}
vis(fit_shan, 'Shannon')
```

```{r}
vis(fit_even, 'Evenness')
```


